---
title: "Trip Distance Calibration"
author: "Mahsa Abdollahyar"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(dplyr)
library(ggplot2)
library(readr)
library(tidyr)
library(stringr)
library(knitr)
library(kableExtra)

trips      <- read_csv("./data/Melbourne/trips.csv")
persons    <- read_csv("./data/Melbourne/households.csv")
households <- read_csv("./data/Melbourne/persons.csv")

```

```{r specifying related variables}

# Prepare categories for each trip
households <- households %>%
  left_join(
    persons %>%
      filter(age >= 18) %>%
      count(hhid, name = "hh_adults"),
    by = "hhid"
  ) %>%
  replace_na(list(hh_adults = 0)) %>%
  mutate(
    carCategory = case_when(
      cars == 0                 ~ "noCar",
      cars / hh_adults < 1      ~ "autosPerAdult_under1",
      TRUE                      ~ "autosPerAdult_over1"
    )
  )

trips <- trips %>%
  left_join(
    households %>% select(hhid, carCategory),
    by = "hhid"
  ) %>% 
  left_join(
    persons %>% select(persid, age, sex, studying, mainact),
    by = "persid"
  ) %>%
  mutate(
    age_group = cut(age,
      breaks = c(0, 17, 29, 59, Inf),
      labels = c("0-17", "18-29", "30-59", "60+"),
      include.lowest = TRUE
    )
  )

# Compute trip lengths
trip <- trips %>%
  mutate(across(dist1:dist9, ~ as.numeric(.x) %>% replace_na(0))) %>%
  rowwise() %>%
  mutate(dist = sum(c_across(dist1:dist9), na.rm = TRUE)) %>%
  ungroup()

# Create clean trips data
trip_final <- trip %>%  
  filter(
    full_purpose %in% c("HBW", "HBE", "HBS", "HBA", "HBO", "HBR", "NHBW", "NHBO"), 
    dist < 25
         ) %>% 
  mutate(employment_status = ifelse(mainact %in% c("Casual Work", "Full-time Work", "Part-time Work"),
                             "Employed", "Unemployed")) %>% 
  select(tripid, purpose, full_purpose, dist, age,
         employment_status,
         student_status = studying,
         car_ownership= carCategory,  
         gender = sex, 
         age_group = age_group)

``` 

# Number of Trips by category

This section presents summary tables showing the total number of trips observed across each category of the selected demographic and socioeconomic variables:

 **Demographic characteristics**   
 - Age group  
 - Gender

**Socioeconomic attributes**  
- Car ownership  
- Student status
- Employment status

```{r table box, results='asis', warning=FALSE, message=FALSE,}

# Table of number of trips for each category
categories <- c("car_ownership", "student_status", "employment_status", "age_group", "gender")

for (cat in categories) {
  cat("\n\n### Number of Trips by ", str_to_sentence(gsub("_", " ", cat)), "\n\n")

  table <- trip_final %>%
    group_by(.data[[cat]]) %>%
    summarise(n_trips = n(), .groups = "drop") %>% 
    rename(
      !!str_to_sentence(gsub("_", " ", cat)) := .data[[cat]],
      `Total trips` = n_trips
    )

  # Print table with controlled width and height
  print(
    table %>%
      kable("html", align = "l") %>%
      kable_styling(
        full_width = FALSE,
        font_size = 18,
        bootstrap_options = c("striped", "hover", "responsive")
      ) %>%
      column_spec(1, width = "20em", bold = TRUE) %>%
      column_spec(2, width = "10em") %>%
      scroll_box(width = "100%", height = "200px")
  )
  
  cat("\n\n<br><br>\n\n")

}

```

# Distribution of Travel Distances by Trip Purpose

This section presents a series of boxplots comparing trip distance distributions across travel purposes for each category of the selected demographic and socioeconomic variables.

```{r generate boxplots, warning=FALSE, message=FALSE, fig.width=12, fig.height=24, out.width='100%'}

# Grid of travel distance boxplots for each category
categories <- c("car_ownership", "student_status", "employment_status", "age_group", "gender")

for (cat in categories) {
  cat("Trip distance by ", gsub("_", " ", cat))
  
  plot <- ggplot(trip_final, aes_string(x = cat, y = "dist")) +
    geom_boxplot(outlier.size = 0.5) +
    facet_wrap(~ full_purpose, ncol = 2, scales = "free_y") +
    labs(
      title = paste0("Trip distance by ", gsub("_", " ", cat)),
      x     = str_to_sentence(gsub("_", " ", cat)),
      y     = "Trip distance"
    ) +
    theme_minimal(base_size = 16) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
  print(plot)
}

```
