---
title: "Micro Destination Attraction Analysis"
format: gfm
execute:
  warning: false
  message: false
---

# Micro Destination Attraction Analysis

This document processes point-of-interest (POI) data for Melbourne to calculate destination attraction coefficients for transport modeling.

## Setup

```{r}
#| label: setup
library(tidyverse)
```

## Load Spatial Zones and Attraction Coefficients

```{r}
#| label: load-zones-attractions
# Read output areas and attraction coefficients
zones <- sf::st_read(
    "../../../../melbourne/input/zonesShapefile/SA1_2016_AUST_MEL.shp"
  ) %>%
  transmute(zone = SA1_MAIN16) %>%
  sf::st_transform(crs = 28355)

attractions <- read_csv("../../../../melbourne/input/mito/tripAttractionsCoefficients.csv")
purposes <- names(attractions)[-1]
```

## Process Destination Data

```{r}
#| label: process-destinations
# all_destinations comes from destinations.r
all_destinations <- read_csv("../../../../melbourne/preprocessing/poi/final/all_destinations.csv")  %>% 
  filter(code != "PT") %>%
  sf::st_as_sf(coords = c("X","Y"),remove = FALSE,crs = 28355) %>%
  sf::st_join(zones,join = sf::st_intersects) %>%
  sf::st_drop_geometry() %>%
  filter(!is.na(zone))  %>%
  mutate(type = recode(type,
    "Community health resources" = "community_health", 
    "Early year access" = "early_year_access",
    "Eating establishments" = "eating_establishments",
    "Education" = "education",
    "Financial" = "financial",
    "Food retail" = "food_retail",
    "Primary health care" = "primary_health_care",
    "Recreational, sports pitches and facilities" = "recreation",
    "Services" = "services",
    "Social and cultural locations" = "social_and_culture",
    "Public open space" = "public_open_space",
  )) %>% 
  rename(WEIGHT = weight) %>%
  group_by(code,id) %>%
  mutate(WT = WT / n()) %>%
  ungroup() %>%
  select(type,X,Y,WEIGHT,Attribute,zone,id,WT,code) %>%
  left_join(attractions, by = c("code" = "poi")) %>%
  mutate(across(all_of(purposes), ~.*WT))
```

## Export Results

```{r}
#| label: export-results
write_csv(all_destinations,"../../../../melbourne/input/mito/microDestinationAttraction.csv")
```

## Data Quality Tests

### Check Uniqueness of Reference Numbers

```{r}
#| label: test-uniqueness
# Check uniqueness of reference numbers
all_destinations %>%
  group_by(type) %>%
  summarise(count = n(),
            dist_ID = n_distinct(id))
```

### Verify ID Uniqueness

```{r}
#| label: verify-ids
# These should match. If not, need to change method for getting reference number
with(all_destinations, n_distinct(code,id))
```

### Weight Distribution by Zone and Code

```{r}
#| label: test-weight-distribution
test <- all_destinations %>%
  group_by(zone,code) %>%
  summarise(sum_wt = sum(WT))
```

## Visualization

### Weight Distribution Over Study Area

```{r}
#| label: plot-weight-distribution
#| fig-cap: "Distribution of weights by POI code across the study area"
plot_data <- all_destinations %>%
  group_by(code) %>%
  summarise(sum_wt = sum(WEIGHT)) %>%
  mutate(p = sum_wt / sum(sum_wt))

ggplot(plot_data, aes(x = p, y = TRUE, fill = code)) + 
  geom_bar(position = "fill", stat = "identity") +
  labs(title = "Weight Distribution by POI Code",
       x = "Proportion",
       y = "",
       fill = "POI Code") +
  theme_minimal()
```

### Count by Zone and Code

```{r}
#| label: count-zone-code
test <- count(all_destinations,zone,code)
```

### Attraction Results by Purpose

```{r}
#| label: plot-attraction-results
#| fig-cap: "Share of attraction by purpose and POI code"
# Plot attraction results
plot_data <- all_destinations %>%
  pivot_longer(cols = all_of(purposes),names_to = "purpose") %>%
  group_by(purpose,code) %>%
  summarise(sum_wt = sum(value)) %>%
  mutate(p = sum_wt / sum(sum_wt))

ggplot(plot_data, aes(x = p,y = purpose, fill = code)) + 
  geom_bar(position = "fill",stat = "identity") + 
  labs(title = "Attraction Share by Purpose and POI Code",
       x = "Share",
       y = "Purpose",
       fill = "POI Code") +
  theme_minimal()
```

### Distribution of Attraction Values

```{r}
#| label: plot-value-distribution
#| fig-cap: "Density distribution of attraction values by purpose"
test <- all_destinations %>%
  pivot_longer(cols = all_of(purposes),names_to = "purpose") %>%
  filter(value > 0) %>%
  ggplot(aes(x = value)) + 
  geom_density() + 
  facet_wrap(~purpose) + 
  xlim(0,0.2) +
  labs(title = "Distribution of Attraction Values by Purpose",
       x = "Attraction Value",
       y = "Density") +
  theme_minimal()

test
```

## Validation

### Test Aggregation

```{r}
#| label: test-aggregation
# Test aggregation before estimating total attraction
# (should match plot data, which estimates attraction before aggregating)
test <- all_destinations %>%
  group_by(code) %>%
  summarise(sum_wt = sum(WT)) %>%
  left_join(attractions, by = c("code" = "poi")) %>%
  mutate(across(all_of(purposes), ~.*sum_wt))

test
```

## Summary

This analysis processes Melbourne POI data to create micro-level destination attraction coefficients for transport modeling. The data is spatially joined with statistical areas, standardised with JIBE naming conventions, and weighted according to attraction coefficients for different trip purposes.
